---
title: "Introduction to Network Concepts"
output: 
  beamer_presentation:
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
library(knitr)
#opts_chunk$set(echo = FALSE)

library(statnet)
library(GGally)
theme_set(theme_minimal() + theme(title=element_text(size = 20)))

```

## Today
- Introduction network terms
- The role of graph theory
- How matrices (ew math) help us

## Nodes and Edges

Networks are made up of nodes that are connected with edges. 

## Network

```{r network, echo=FALSE}

set.seed(1)
net <- rgraph(n=10, m=1, tprob=.2, mode='digraph')
colnames(net) <- LETTERS[1:10]
net <- network(net)
ggnet2(net, mode="circle", color='orangered3', arrow.size = 10,  arrow.gap = 0.025)

```


## Network

```{r network_node, echo=FALSE}
ggnet2(net, mode="circle", color='orangered3', arrow.size = 10,  arrow.gap = 0.025) + 
  annotate("segment", x=.8, y=.34, xend=.97, yend=.34, arrow=arrow()) + 
  annotate("text", x=.8, y=.34, label="Node", hjust=1, size=10)
```


## Network

```{r network_edge, echo=FALSE}

ggnet2(net, mode="circle", color='orangered3', arrow.size = 10,  arrow.gap = 0.025) + 
  annotate("segment", x=.8, y=.34, xend=.97, yend=.34, arrow=arrow()) + 
  annotate("text", x=.8, y=.34, label="Node", hjust=1, size=10) + 
  annotate("segment", x=.65, y=.45, xend=.8, yend=.45, arrow=arrow()) + 
  annotate("text", x=.65, y=.45, label="Edge", size=10, hjust=1)
```

## Graph Theory

This makes sense visually, but in order to analyze it we need to use *math*

## Graph theory

Graph:

> - Is defined as a set of vertices (nodes) and edges: G(V, E)
> - Start with a set of vertices: V={A, B, C, D}
> - Edges connect two vertices, making them adjacent: (A, C)
> - E then the set of such connections: E={(A,C), (A,D), (B,C)} 
  >   - Note: these edges have a direction: (A, C), is different from (C,A)

## Network as a Graph

```{r network_labeled, echo=FALSE, results='asis', out.height='60%' ,fig.align='center'}
tmp <- apply(as.edgelist(net), 1, function(x) paste0("(", LETTERS[x[1]], ",", LETTERS[x[2]], ")"))
cat(paste0("E={",paste0(tmp, collapse = ", "),"}"))
cat("\n")
ggnet2(net, mode="circle", color='orangered3', arrow.size = 10,  
       arrow.gap = 0.05, label=TRUE, size=20, label.size=15)
```
## Adjacent, Incident, and Degree

- Two vertices that are connected by an edge are **adjacent**
- An edge that is connected to a vertice is said to be **incident** to it. 
- The number of edges **incident** to a vertex is its **degree** 

## Degree

A has 4 edges incident to it, so it has degree of 4
```{r network_deg, echo=FALSE, out.height='60%' ,fig.align='center'}
net %e% "color" <- "gray"
set.edge.attribute(net, "color",value='springgreen4',
                   e=get.edgeIDs(net, 1, neighborhood = "combined"))
ggnet2(net, mode="circle", color='orangered3', arrow.size = 10,  edge.size = 2,
       arrow.gap = 0.05, label=TRUE, size=20, label.size=15, edge.color = "color")
```
## Degree for E


What is E's degree? 

```{r network_deg2, echo=FALSE, out.height='60%', fig.align='center'}
net %e% "color" <- "gray"
ggnet2(net, mode="circle", color='orangered3', arrow.size = 10,  edge.size = 2,
       arrow.gap = 0.05, label=TRUE, size=20, label.size=15, edge.color = "color")
```

## Degree for E

```{r network_deg3, echo=FALSE, out.height='60%', fig.align='center'}
set.edge.attribute(net, "color",value='springgreen4',
                   e=get.edgeIDs(net, 5, neighborhood = "combined"))
ggnet2(net, mode="circle", color='orangered3', arrow.size = 10,  edge.size = 2,
       arrow.gap = 0.05, label=TRUE, size=20, label.size=15, edge.color = "color")
```


## Undirected networks

It is also possible to have networks where edges do not have a direction: 

```{r network_un, echo=FALSE, out.height='60%',fig.align='center', results='asis'}

set.seed(1)
net <- rgraph(n=10, m=1, tprob=.2, mode='graph')
colnames(net) <- LETTERS[1:10]
net <- network(net,directed = F)
tmp <- apply(as.edgelist(net), 1, function(x) paste0("(", LETTERS[x[1]], ",", LETTERS[x[2]], ")"))
cat(paste0("E={",paste0(tmp, collapse = ", "),"}"))
cat("\n")
ggnet2(net, mode="circle", color='orangered3', label=TRUE, size=20, label.size=15)
```


## Directed vs Undirected Network

Whether a network is directed or undirected is dependent on the types of relationships:

### Directed: 
 - Parent-Child
 - Donor-Donee

### Undirected
 - Friend (hopefully)
 - Classmate 
 - Ally 

## Characteristics for nodes

What are some things that might matter if you are at a node? 
- Where can you move to?
- How fast can you get there? 
- Who can you reach? 
  
## Moving along a network 
We can move from one vertex (node) to another through edges.

- A walk is a sequence of **adjacent** vertices: 
  - G-A-D-H
  - F-C-E-F-I
  - H-D-A
- A path is a sequnce of **adjacent** vertices where none are repeated:
  - G-A-D-H
  - H-D-A
- All paths are walks, but not all walks are paths. 

## The length of paths and walks

- The **length** of a walk or paths is how many edges are in them. 
- The shortest path is called a **geodesic**.
- The **(geodesic) distance** between two vertices is the length their **geodesic**


## Geodesic Distance

The distance between F and C is 1
```{r network_dist1, echo=FALSE, out.height='60%',fig.align='center', results='asis'}
net %e% "color" <- "gray"
set.edge.attribute(net, "color", "springgreen4", 
                   e=c(get.edgeIDs(net, 6, 3)))

ggnet2(net, mode="circle", color='orangered3',  edge.size = 2,
       label=TRUE, size=20, label.size=15, edge.color = "color")
```


## Geodesic Distance

The distance between G and C is 4
```{r network_dist2, echo=FALSE, out.height='60%',fig.align='center', results='asis'}
net %e% "color" <- "gray"
set.edge.attribute(net, "color", "springgreen4", 
                   e=c(get.edgeIDs(net, 7, 1), get.edgeIDs(net, 1, 4),
                       get.edgeIDs(net, 4, 6), get.edgeIDs(net, 6, 3)))

ggnet2(net, mode="circle", color='orangered3',  edge.size = 2,
       label=TRUE, size=20, label.size=15, edge.color = "color")
```

## Why care about distance?

- Distance is viewed as how 'close' two vertices are. 
- Information spreads faster between two vertices that are closer

## Looking at some real data: Twitter Senate Data

\begin{center}

Twitter follow network among US Senators

\end{center}

```{r, echo=FALSE, out.height='60%',fig.align='center', results='hide'}
adj_df <- read.csv("data/senate_adj.csv")
meta_df <- read.csv("data/senate_data.csv")
adj_df$X <- NULL
adj_df <- as.matrix(adj_df)
adj_df <- adj_df=="True"

present <- nchar(colnames(adj_df)) != 3

colnames(adj_df) <- meta_df$name
rownames(adj_df) <- meta_df$name[present]
adj_df <- adj_df[,present]
meta_df <- meta_df[present, ]


sen_net <- network(adj_df, directed = T)
sen_net %v% "party" <- ifelse(meta_df$party=="democrat", 
                              "skyblue2", ifelse(meta_df$party=="republican",
                                                 "orangered3", 
                                                 "springgreen3"))

delete.vertices(sen_net, isolates(sen_net))
set.seed(1)
ggnet2(sen_net,mode = "fruchtermanreingold",
       color='party',  edge.size = .5,
       edge.alpha = .5,
       label=F, size=5, label.size=10, 
        arrow.size = 3,  
       arrow.gap = 0.02 ) + guides(color=F)
```

## Looking at some real data

```{r, echo=FALSE, out.height='60%',fig.align='center', results='hide'}

dists <- geodist(sen_net, predecessors = T, inf.replace = NA)

opts <- which(dists$counts==1 & dists$gdist>2, arr.ind = T)

names <- sen_net %v% "vertex.names"
names <- sapply(names, function(x) strsplit(x, " ")[[1]][2])

dists$gdist[opts[1,1],opts[1,2]]
dists$counts[opts[1,1],opts[1,2]]

e_list <- numeric(dists$gdist[opts[1,1],opts[1,2]])
name_list <- numeric(length(e_list) + 1)
end <- opts[1,2]
name_list[1] <- names[end]

for(ii in 1:dists$gdist[opts[1,1],opts[1,2]]){
  tmp <- dists$predecessors[[opts[1,1]]][[end]]

  e_list[ii] <- get.edgeIDs(sen_net, tmp, alter=end)
  end <- tmp 
  name_list[ii + 1] <- names[end]
}

sen_net %e% "color" <- "gray"
sen_net %e% "size" <- .5
sen_net %v% "name" <- ""

set.vertex.attribute(sen_net, "name", names[opts[1,]],v = opts[1,])
set.edge.attribute(sen_net, "color",value = "deeppink1", e = e_list)
set.edge.attribute(sen_net, "size",value = 2, e = e_list)

set.seed(1)
ggnet2(sen_net,mode = "fruchtermanreingold",
       color='party',  edge.size ="size",
       edge.alpha = .5, edge.color='color',
       label="name", size=5, label.size=10, 
        arrow.size = 3,  
       arrow.gap = 0.02 ) + guides(color=F)
```


```{r, echo=FALSE, results='asis'}
name_list <- rev(name_list)
cat(paste(name_list[1], "follows", paste0(name_list[-1], collapse=", who follows ")))
```


## Looking at some real data

```{r, echo=FALSE, out.height='60%',fig.align='center', results='hide'}
e_list <- numeric(dists$gdist[opts[1,2],opts[1,1]])

name_list <- numeric(length(e_list) + 1)
end <- opts[1,1]

name_list[1] <- names[end]

for(ii in 1:dists$gdist[opts[1,2],opts[1,1]]){
  tmp <- dists$predecessors[[opts[1,2]]][[end]][1]

  e_list[ii] <- get.edgeIDs(sen_net, tmp, alter=end)
  end <- tmp 
  name_list[ii + 1] <- names[end]

}


set.edge.attribute(sen_net, "color",value = "deeppink1", e = e_list)
set.edge.attribute(sen_net, "size",value = 2, e = e_list)

set.seed(1)
ggnet2(sen_net,mode = "fruchtermanreingold",
       color='party',  edge.size ="size",
       edge.alpha = .5, edge.color='color',
       label="name", size=5, label.size=10, 
        arrow.size = 3,  
       arrow.gap = 0.02 ) + guides(color=F)
```

```{r, echo=FALSE, results='asis'}

name_list <- rev(name_list)
cat(paste(name_list[1], "follows", paste0(name_list[-1], collapse=", who follows ")))
```


## What can we learn?

- The mean distance between any two senators is `r round(mean(dists$gdist, na.rm=T), 2)`
- The median distance between any two senators is `r median(dists$gdist, na.rm=T)`
- The median degree is `r median(degree(sen_net))`


## Degree for Directed Networks

Directed networks have both **outdegree** (how many edges start at a vertex) and **indegree** (how many edges point at a vertex):

- Median outdegree (how many each Senator follows): `r median(degree(sen_net, cmode="outdegree"))`
- Median indegree (how many followers each Senator has): `r median(degree(sen_net, cmode="indegree"))`

## Who is the most popular? 
 
```{r, echo=FALSE, results='asis'}
names <- sen_net %v% "vertex.names"

degs_out <- degree(sen_net, cmode='outdegree')
degs_in <- degree(sen_net, cmode='indegree')

degs <- data.frame("Following"=degs_out, "Followers"=degs_in)
rownames(degs) <- names
kable(degs[1:5,])
```

- `r names[which.max(degs_out)]` follows the most (follows `r max(degs_out)`)
- `r names[which.max(degs_in)]` is followed by the most (follows `r max(degs_in)`)


## Who can reach who?

- A vertex with no incident edges is known as an **isolate**
- We can also group vertices by who they can reach. 

\begin{center}
\vspace{1em}
A \textbf{component} is the maximal (largest) group of vertices where every vertex within it can reach every other vertex.

\end{center} 

## Some Components


```{r network_comp, echo=FALSE, out.height='60%',fig.align='center', results='hide'}
set.seed(1)
n <- 20
net <- rgraph(n=n, m=1, tprob=.1, mode='graph')
colnames(net) <- LETTERS[1:n]
net <- network(net,directed = F)
comp <- component.dist(net)
net %v% "component" <- comp$membership
ggnet2(net, color='component',  edge.size = 2,
       label=TRUE, size=20, label.size=15, palette = "Set2") + guides(color=F)

```
\begin{center}

This network has 4 components, the largest has 15 vertices in it. 
\end{center}

## Components with Directed Graphs

When a network is directed, then we need to think about direction.

- **Weak Component:** Is a component if we disregard direction of edges.
- **Strong Component:** Is a component if we follow direction of edges

## Strong/Weak Components:


```{r network_comp_dir, echo=FALSE, out.height='60%',fig.align='center', results='hide'}
set.seed(711)
n <- 20
net <- rgraph(n=n, m=1, tprob=.09, mode='digraph')
colnames(net) <- LETTERS[1:n]
net <- network(net,directed = T)
comp <- component.dist(net, connected="strong")
net %v% "component" <- comp$membership
p1 <- ggnet2(net, color='component',  edge.size = 2,
       label=TRUE, size=15, label.size=10, palette = "Set2",        
       arrow.size = 10,  
       arrow.gap = 0.05) + guides(color=F) + ggtitle("Strong Component")

lo <- p1$data[,c("x","y")]
comp <- component.dist(net,connected = "weak")
net %v% "component" <- comp$membership
p2 <- ggnet2(net, mode=as.matrix(lo),
       color='component',  edge.size = 2,
       label=TRUE, size=15, label.size=10, palette = "Set2", 
        arrow.size = 10,  
       arrow.gap = 0.05 ) + guides(color=F) +ggtitle("Weak Component")
gridExtra::grid.arrange(p1, p2,ncol=2)
```


## Real Life Data

```{r, echo=FALSE, out.height='60%',fig.align='center', results='hide'}
comp <- component.dist(sen_net, connected="strong")
sen_net %v% "component" <- comp$membership
set.seed(1)
p1 <- ggnet2(sen_net, color='component', edge.size = .5,
        edge.alpha = .5,
        size=5, label.size=10, 
        arrow.size = 3,
       label=F, palette = "Set2", 
       arrow.gap = 0.02) + guides(color=F) + ggtitle("Strong Component")

lo <- p1$data[,c("x","y")]
comp <- component.dist(sen_net,connected = "weak")
sen_net %v% "component" <- comp$membership
p2 <- ggnet2(sen_net, mode=as.matrix(lo),
       color='component',  edge.size = .5,
        edge.alpha = .5,
        size=5, label.size=10, 
        arrow.size = 3,
       label=F, palette = "Set2", 
       arrow.gap = 0.02 ) + guides(color=F) +ggtitle("Weak Component")
gridExtra::grid.arrange(p1, p2,ncol=2)
```


## Bring the math back in

How do we organize this data? An **adjacency matrix**

## Matrix Referesher 

$$\mathbf{A} = \left[\begin{array}
{rrr}
0 & 0 & 0 \\
0 & 0 & 1 \\
0 & 0 & 0
\end{array}\right]$$

- A matrix is a rectangular collection of numbers (n $\times$ m in size). 
- Each number (element) is given a position by what row and column it is. 
- The 1 above is in row 2, column 3. 

## Adjacency matrix

- For now our matrices will be square: n $\times$ n where n is the number of vertices. 
- A 1 means an edge exists from that row to that column, a 0 means no edge exists. 

\pause 
$$\mathbf{A} = \left[\begin{array}
{rrr}
0 & 0 & 1 \\
0 & 0 & 0 \\
0 & 1 & 0
\end{array}\right]$$

- In **A** there is an edge from node 3 to node 2, and from node 1 to node 3 (directed). 

## Adjacency Matrix vs Network Graph

```{r, echo=FALSE, out.width='45%', fig.align='center'}
set.seed(1)
n <- 6
mat <- rgraph(n=n, m=1, mode='graph')
colnames(mat) <- LETTERS[1:n]
rownames(mat) <- LETTERS[1:n]
net <- network(mat,directed = F)
ggnet2(net,  edge.size = 2, mode='circle',
       label=TRUE, size=25, label.size=20)
```

```{r, echo=FALSE, results='asis', out.width='45%'}
print(xtable::xtable(mat,align="lcccccc", digits=0), comment=F)
```

## Undirected Adjacency Matrix

- For undirected graphs the adjacency matrix is mirrored across the diagonal
- Meaning that the entry for  $a_{i,j}$ is the same as for  $a_{j,i}$

## Senator Adjacency Matrix 

```{r, echo=FALSE,results="asis"}
adj_df <- 1*adj_df
adj_mat <- adj_df[1:5,1:3]
rownames(adj_mat)[3] <- "Doug Jones"
colnames(adj_mat)[3] <- "Doug Jones"

print(xtable::xtable(adj_mat,align="lccc",digits=0), comment= F)

```