---
title: "Network Research Design and Data Collection"
author: "Kevin Reuning"
output: 
  beamer_presentation:
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
library(knitr)
#opts_chunk$set(echo = FALSE)

library(statnet)
library(GGally)
theme_set(theme_minimal() + theme(title=element_text(size = 20)))

```



## Goals for Today 

- Refresher on experimental, quasi-experimental and observational designs
- Whole-network versus ego-networks
- Gathering data on relationships

## Research Designs

Three broad types of research designs based on how the independent variable is set:

- **Experimental Design:** The researcher assigns the level of the independent variable (usually randomly) to each unit
- **Quasi-Experimental Design:** The researcher makes an argument that the level of the independent variable is assigned as if randomly by nature. 
- **Observational:** No claims of random assignment of independent variable level, attempts to control for confounders in other ways. 

## (Quasi-)Experimental Designs in Networks

- Less common in social network analysis 
- Researcher can manipulate networks in lab conditions, but outside of labs it becomes hard. 
- Often the network is not directly manipulated but is part of an experiment in other ways. 


## Observational Designs 

- Much more common in social network analysis. 
- Need to collect network data: ego-network or whole person network.

## Network Research Examples


## Whole vs Ego(Person)-Network 

Two ways to approach network data: 

- Whole Network: Collect the 'entire' network.
- Ego-Network: Collect each individual's immediate network.

## Whole Network Approach

What do you need to do this? 

1. Ability to identify every node. 
2. Ability to identify all relationships between nodes. 


## Whole Network: Bounding

Identifying relevant nodes should be informed from research question. 

\pause 

- **Question**: Are interest groups lobbying relationships explained by overlap in donors?
- **Bound:** All interest groups actively lobbying the federal government in a given year. 

\pause 

- **Question:** Do popular people at protests tend to hold more extreme views or more moderate views? 
- **Bound:** All individuals who participate at a specific protest.

## Whole Network: Bounding

You need to be able to explain why some potential nodes are part of or not part of your research. 

\pause 

- **Good reasons:** Not part of a natural group or outside the scope of research. 
- **Bad reasons:** Node does not have any relationships (no relationships is interesting!) or it would be too hard to identify their relationships.


## Whole Network: Identifying Relationships

Your research question should imply a type of relationship of interest. 

\pause 
 
Remember we have four types of relationships: 
- Similarities
- Relations
- Interactions
- Flows

## Whole Network: Identifying Relationships 


- **Question**: Are interest groups lobbying relationships explained by overlap in donors?
- **Relationship Options:** 
  - Are they both lobbying on same side of the same issue? _(similarity)_
  - Are they meeting with each other? _(interaction)_
  - Is information passing between them? _(flow)_

\pause 

- **Question:** Do popular people at protests tend to hold more extreme views or more moderate views? 
- **Relationship Options:** 
  - Do they consider a person a friend? _(relationship)_
  - Did they talk with each other? _(interaction)_
  - Did they stand near each other? _(similarity)_
  
## Whole Network: Identifying Relationships

We often have to infer relationships from something that isn't exactly what we are interested in. 

\pause 

- **Question:** Do popular people at protests tend to hold more extreme views or more moderate views? 
- **Relationship Options:** 
  - Do they consider a person a friend? **Hard to do even without a lot of people**
  - Did they talk with each other? **Feasible to do if the crowd isn't too big**
  - Did they stand near each other? **Feasible to do if you can identify individuals**

\pause 

We might use 'talking to' and 'standing near' even though we really want friendships. 


## Ego-Network 

Sometimes it isn't reasonable to bound a network. Instead you focus on an ego-network:

- Ego-networks just include who your node is connected to (i.e. list all your friends)
- Can still do some interesting things, but not as much. 

\pause 

\begin{center}
Most of the time we will focus on whole network research. 
\end{center}

## Collecting Data

Types of data collection: 
- Survey data (often, but not necessarily for ego-networks)
- Behavioral data (often, but not necessarily for whole networks)

## Survey data

You can collect networks by just asking people! But...

- You need to think through wording: What does friend mean? What does colleague mean?
- Are you going to give someone a list of names or just let them fill in names? 
- Who really wants to list all their friends? 

## Survey Data: Ways to Improve Question Wording

- Provide as little ambiguity in language as possible. 
- Try not to make assumptions about individual behavior. 
- Before you do a survey do a focus group with potential participants. 

## Survey Data: Ways to Improve Question Wording

What if you are interested in identifying friends within a college? 

**Bad:**
- Who do you consider a friend? 
- Who do you hangout with often?
- Who do you go to a bar with? 

**Good:**
- In a normal week who do you spend time with outside of class or work? 
- Who do you ask for advice from? 

## Survey Data: Ways to Improve Format

Questions can be either be open-ended or closed-ended

- Open-ended: Individuals write out their responses.
  - Benefits: You do not need a list of potential names. 
  - Costs: Can be hard (impossible) to match across participants and people are forgetful. 
- Closed-ended: Individuals are given a list to pick from.
  - Benefits: You can match across across participants and recall is improved.
  - Costs: You need a list of potential names. 

## Survey Data: Ways to Improve Format

If you are trying to do a whole network survey you need to do closed-ended, if you are doing an ego-network you can do open-ended. 

## Behaviorial Data: 

There are a lot of other sources of networks:
- Social media and the internet
- Archival sources (collections of letters for example)
- Organizational memberships 

## Behaviorial Data Benefits and Costs

Benefits: 
- You can often get something close to a whole network. 
- Can relatively easily collect data overtime (watch network evolve)
- Often is already out there for you download

Costs: 
- It is not always clear what relationships might mean. 
- You have to deal with gatekeepers who might not want to share data. 

## Behaviorial Data Example


\begin{center}

Twitter follow network among US Senators

\end{center}

```{r, echo=FALSE, out.height='60%',fig.align='center', results='hide'}
adj_df <- read.csv("data/senate_adj.csv")
meta_df <- read.csv("data/senate_data.csv")
adj_df$X <- NULL
adj_df <- as.matrix(adj_df)
adj_df <- adj_df=="True"

present <- nchar(colnames(adj_df)) != 3

colnames(adj_df) <- meta_df$name
rownames(adj_df) <- meta_df$name[present]
adj_df <- adj_df[,present]
meta_df <- meta_df[present, ]


sen_net <- network(adj_df, directed = T)
sen_net %v% "party" <- ifelse(meta_df$party=="democrat", 
                              "skyblue2", ifelse(meta_df$party=="republican",
                                                 "orangered3", 
                                                 "springgreen3"))

delete.vertices(sen_net, isolates(sen_net))
set.seed(1)
ggnet2(sen_net,mode = "fruchtermanreingold",
       color='party',  edge.size = .5,
       edge.alpha = .5,
       label=F, size=5, label.size=10, 
        arrow.size = 3,  
       arrow.gap = 0.02 ) + guides(color=F)
```

## What does a follow mean? 

What does it mean when a senator follows another senator? 

## Today

- Social network research is often observational given limitations. 
- Whole-network is about collecting the...whole... network, ego-network is about an individual's personal network. Both can work, but you need to think about your research question. 
- Surveying network relationships is hard, collecting behaviorial data can be a lot easier but requires interpretation. 




